---
title: "Trends in Heating and Cooling Days in Suburban and Urban Settlements in the U.S. Southwest"
author: "Candela Cerpa & Samantha Pace & Vinny Whatley"
date: "2023-12-13"
output: pdf_document
fig_caption: yes
---
\newpage
\tableofcontents 
\newpage
\listoftables 
\listoffigures 
\newpage

```{r setup, include=FALSE}

# Load your packages
#install.packages("ggplot2")
#install.packages("ggthemes")
#install.packages("cowplot")
#install.packages("knitr")
#install.packages("Kendall")

library(tidyverse)
library(lubridate)
library(here)
library(ggplot2)
library(ggthemes)
library(cowplot)
library(zoo)
library(knitr)
library(Kendall)

# Set your working directory
here()
```

```{r setting global code chunks, include =FALSE}
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
               fig.height = 3, fig.width = 6)
```

```{r set ggplot theme}
# Set your ggplot theme
theme <- theme_base() +
  theme(
   plot.background = element_rect(
      color = "black",
      fill = "gray"
    ),
    legend.position = "top", 
   legend.direction = "horizontal",
    complete = TRUE
  )
theme_set(theme)

#monthly labels
month_lab = tibble(doy = yday(as.Date(c("1980-02-15", "1980-04-15",
                                          "1980-06-15", "1980-08-15",
                                          "1980-10-15", "1980-12-15"))),
                     lab = c("Feb", "Apr", "Jun", "Aug", "Oct", "Dec"))
```

# 1 Rationale and Research Questions
Climate change, long-term shifts in temperatures and weather patterns, is a naturally-occurring process that has been accelerated by human actions, primarily the burning of fossil fuels. One of the impacts of climate change is rising temperatures, which has negative impacts on Earth's ecosystems, such as melting glaciers, intensifying storms, and worsening drought conditions. Humans are directly impacted by rising temperatures: as the number of hot days and heat waves increase, so do the number of heat-related illnesses.

The U.S. has not escaped these impact. The southwest region of the US which we defined as Arizona, California, and Nevada has witnessed an uptick in droughts. Nevada which utilizes water from Lake Me has seen record lows in recent years cautioning them to adopt water conservation strategies while California has been no stranger to forest fires. We plan to do this by analyzing data from the National Centers for Environmental Information, using max temperature and minimum temperature.  Thus, our research questions are:

1. Has there been an increase in the number of heating and cooling days in the U.S. Southwest from 1980 to 2022?

2. Is there a difference in heating and cooling days in urban and rural spaces in the U.S. Southwest over time?

3. How have minimum and maximum temperatures changed over time?

# 2 Dataset Information
Our datasets were retrieved from NOAA's National Centers for Environmental Information (NCEI), specifically from the GHCN (Global Historical Climatology Network). We selected a total of six stations, encompassing three major cities and each with an associated nearby suburban location. We  selected the nearby suburban locations to be within 50 miles of the major city and with a population of less than 50,000 in 2022. The selected locations are as follows:

* Phoenix and Fountain Hills, Arizona
* Las Vegas and Pahrump, Nevada
* San Diego and Ramona, California

We primarily examined the daily data for TMAX = Maximum temperature (Fahrenheit) and TMIN = Minimum temperature (Fahrenheit) for each of the six locations. Our datasets had varying levels of coverage and start dates for each location. All the datasets had high coverage from 1998-2022, and all but Ramona, CA have data from 1980-1998. More information about this data's documentation can be found at https://www.ncei.noaa.gov/pub/data/cdo/documentation/GHCND_documentation.pdf and in the Metadata folder of this project's repository.

\newpage

# 3 Data Wrangling

We took several steps in our workflow to clean our datasets. The raw data included both the city and suburb in the same CSV, so we first filtered the data to separate out the locations. For each of the 6 locations, we selected the columns we were interested in exploring, which included NAME, DATE, PRCP, PSUN, TMAX, TMIN, TAVG, TSUN, and created columns for Month, Year, and DayofYear to enable different years to be overlaid on a graph.

The Fountain Hills, Pahrump, and Ramona datasets had some missing data points in the maximum and minimum daily temperatures, so we used linear interpolation to fill in the missing data. The raw data for the 'Average Temperature' column contained many missing values, so we inputted the calculated mean of the daily maximum and minimum for this column.

## Heating and Cooling Degree Days Calculations

Having the daily average temperature calculated for each location, we then calculated the heating degree days and cooling degree days. Heating degree days are when people turn their heater on because the average temperature is low, and cooling degree days is when people turn their air conditioner on because the average temperature is warm. The industry standard threshold is 65 degrees F, meaning that anything higher than 65 F would constitute a cooling degree day, and anything lower than 65 would mean it is a heating degree day (US EPA, 2016). The equations to calculate the heating and cooling degree days are as follows:

Heating degree days = 65 - daily average temperature

Cooling degree days = daily average temperature - 65

From these calculations we are able to look at the number of heating and cooling degree days, as well as the intensity, which are two main components examined in the exploratory analysis section.

*Describe how you wrangled your dataset in a format similar to a methods section of a journal article.*

To wrangle our data, we:
* Read in the .csv’s
* Filtered by location and created new files for each location
* Kept certain columns (name, date, precipitation, sun coverage %, max temp, min temp, average tem, total sun)
* Format the date column into a date object
* Changed the station name to the city name
* Created a column for month, year, and day of year by extracting the information from the existing date column
* Updated the column names to the full title
** Check the NAs with the summary function
* Saw that some cities have NAs in max, min, and average temperature
* Interpolate gaps in min and max using linear interpolation in Fountain Hills, Pahrump, and Ramona
* Calculate the average daily temperature and fill in the existing column
* Created separate columns for heating and cooling degrees and calculated the values
** As we created the two columns, we filled them with 0s
** We calculate heating degrees by indexing the mean temperatures that are below 65F then subtracting the mean temp from 65
** We calculate cooling degrees by indexing the mean temperatures that are above 65F then subtracting 65 from the mean temp

*Add a table that summarizes your data structure (variables, units, ranges and/or central tendencies, data source if multiple are used, etc.). This table can be made in markdown text or inserted as a kable function in an R chunk. If the latter, do not include the code used to generate your table.*

\newpage

# 3. Exploratory Analysis 

For the purposes of exploring our data, we first plotted the three major cities together to get a sense of how their average daily temperatures compared over time, shown in Figure 1. Based on the figure, there are clear seasonal trends throughout the year, and Phoenix has the highest daily average temperature over time, followed by Las Vegas and then San Diego. The trendline for Las Vegas has a noticeable positive slope over time, while the trendline for Phoenix demonstrates only a slight increase and San Diego's trendline remains to be horizontal with no major changes in slope.

```{r Load Processed files}
Phoenix <- read.csv(here("./Data/Processed/PhoenixAZ_processed.csv"),
                    stringsAsFactors = TRUE) %>% 
  mutate(Date = ymd(Date))

FountainHills <- read.csv(here("./Data/Processed/FountainHillsAZ_processed.csv"),
                          stringsAsFactors = TRUE) %>% 
  mutate(Date = ymd(Date))

LasVegas <- read.csv(here("./Data/Processed/LasVegasNV_processed.csv"),
                     stringsAsFactors = TRUE) %>% 
  mutate(Date = ymd(Date))

Pahrump <- read.csv(here("./Data/Processed/PahrumpNV_processed.csv"),
                    stringsAsFactors = TRUE) %>% 
  mutate(Date = ymd(Date))

SanDiego <- read.csv(here("./Data/Processed/SanDiegoCA_processed.csv"),
                     stringsAsFactors = TRUE) %>%
  mutate(Date = ymd(Date))

Ramona <- read.csv(here("./Data/Processed/RamonaCA_processed.csv"),
                   stringsAsFactors = TRUE) %>% 
  mutate(Date = ymd(Date))
```

```{summary table, eval=FALSE, include=FALSE}
#sum_city <- c("Phoenix", "Fountain Hills", "Las Vegas", "Pahrump", "San Diego", "Ramona")
#sum_date <- c("1980-01-01 to 2022-12-31", "1980-01-01 to 2022-12-31", "1972-01-01 to 2022-12-31", "1972-01-01 to 2022-12-31", "1972-01-01 to 2023-01-01", "1998-04-16 to 2023-01-01")
#sum_precip <- c()
#sum_perc_sun <- c()
#sum_daily_temp_max <- c()
#sum_daily_temp_min <- c()
#sum_temp_avg <- c()
#sum_total_sun <- c()
#sum_month <- c("1 to 12", "1 to 12", "1 to 12", "1 to 12", "1 to 12", "1 to 12")
#sum_year <- c("1980 to 2022", "1980 to 2022", "1972 to 2022", "1972 to 2022", "1972 to 2023", "1998 to 2023")
#sum_dayofyear <- c("1 to 366", "1 to 366", "1 to 366", "1 to 366", "1 to 366", "1 to 366")
#sum_heat_degrees <- c("0 to ", "0 to ", "0 to ", "0 to ", "0 to ", "0 to ")
#sum_cool_degrees <- c("0 to ", "0 to ", "0 to ", "0 to ", "0 to ", "0 to ")

#summary_table <- data.frame(sum_city, sum_date, sum_precip, sum_perc_sun, sum_daily_temp_max, sum_daily_temp_min, sum_temp_avg, sum_total_sun, sum_month, sum_year, sum_dayofyear, sum_heat_degrees, sum_cool_degrees)

#kable(summary_table, "pipe", col.names =  c("Location", "Date", "Precipitation", "Percent_Sunshine_Possible",
    "Daily_Temp_Max", "Daily_Temp_Min", "Temp_Avg", "Total_Sunshine",
    "Month", "Year", "DayofYear"), 
#caption = "Data Structure Summary", align = c("c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c"))
```

```{r all cities avg temp, fig.cap= "Average Tempearture for Phoenix, Las Vegas, San Diego", fig.height = 2.75}
Three_cities <- rbind(LasVegas,
                      Phoenix,
                      SanDiego)

ggplot(Three_cities, aes(x = Date,
                         y = Temp_Avg,
                         color = Location, 
                         alpha = 0.75)) +
  geom_line() +
  geom_smooth(method=lm) + 
  labs(title = "Three Cities' Daily Average Temperatures",
       y = "Daily Avg Temp in degrees F") 
```

For the remainder of the exploratory analysis section, we focused on a single city - Phoenix, Arizona - in order to better understand the data for the statistical analyses. Figure 2 shows Phoenix's maximum and minimum daily temperature from 1980 until 2022. The trend for the minimum and maximum temperatures show that there has been a modest increase in both over time. The black horizontal line on Figure 2 marks the 65 F threshold for heating and cooling degrees. Based on the trends in this figure, we would expect there to be more cooling degree days than heating.

```{r plots exploring the trends in the min and max temperatures, fig.cap="Phoenix Daily Maiximum and Minimum Temperatures", fig.height = 2.75}
# Creating a plot with Phoenix's min and max temp
ggplot(Phoenix, aes(x = Date)) +
     	geom_line(aes(y = Daily_Temp_Max, color = "Max Temp")) +
  geom_smooth (aes(y = Daily_Temp_Max), method = lm, color = "red")+
  geom_line(aes(y = Daily_Temp_Min, color = "Min Temp")) +
  geom_smooth(aes(y = Daily_Temp_Min), method = lm, color = "blue") +
  labs( title = "Phoenix Daily Maximum and Minimum Temperatures 1980-2022", y = "Temperature in F") +
  geom_hline(yintercept = 65)
```

```{r code to plot number of heating and cooling days}
#Create dataframes for heating and cooling days, separately
#Make one for 1980 and 2022 to compare the start and end of the time series
Phoenix_Heat_1980 <- Phoenix %>%
  group_by(Heat_Degrees) %>%
  filter(Year == "1980", Heat_Degrees != 0) %>%
  select(Date, Month, Heat_Degrees)

Phoenix_Cool_1980 <- Phoenix %>%
  group_by(Cool_Degrees) %>%
  filter(Year == "1980", Cool_Degrees != 0) %>%
  select(Date, Month, Cool_Degrees)

Phoenix_Heat_2022 <- Phoenix %>%
  group_by(Heat_Degrees) %>%
  filter(Year == "2022", Heat_Degrees != 0) %>%
  select(Date, Month, Heat_Degrees)

Phoenix_Cool_2022 <- Phoenix %>%
  group_by(Cool_Degrees) %>%
  filter(Year == "2022", Cool_Degrees != 0) %>%
  select(Date, Month, Cool_Degrees)

#Plot heating days and cooling days in separate graph bars
plot_Phoenix_1980_Heat <- ggplot(Phoenix_Heat_1980, aes(x = Month)) +
  geom_bar(fill = "red") +
  scale_x_continuous(breaks = seq_along(month.abb), labels = month.abb) +
  labs(title = "Heating Days in Phoenix, 1980", x = "Month", y = "Heating Days")

plot_Phoenix_1980_Cool <- ggplot(Phoenix_Cool_1980, aes(x = Month)) +
  geom_bar(fill = "blue") +
  scale_x_continuous(breaks = seq_along(month.abb), labels = month.abb) +
  labs(title = "Cooling Days in Phoenix, 1980", x = "Month", y = "Cooling Days")

grid_Phoenix_1980_HeatCoolDays <- plot_grid(
  plot_Phoenix_1980_Cool, plot_Phoenix_1980_Heat)

#2022 plots
plot_Phoenix_2022_Heat <- ggplot(Phoenix_Heat_2022, aes(x = Month)) +
  geom_bar(fill = "red") +
  scale_x_continuous(breaks = seq_along(month.abb), labels = month.abb) +
  labs(title = "Heating Days in Phoenix, 2022", x = "Month", y = "Heating Days")

plot_Phoenix_2022_Cool <- ggplot(Phoenix_Cool_2022, aes(x = Month)) +
  geom_bar(fill = "blue") +
  scale_x_continuous(breaks = seq_along(month.abb), labels = month.abb) +
  labs(title = "Cooling Days in Phoenix, 2022", x = "Month", y = "Cooling Days")

# grid_Phoenix_2022_HeatCoolDays <- plot_grid(
#  plot_Phoenix_2022_Cool, plot_Phoenix_2022_Heat)
```

\newpage

Figures 3 and 4 compare the number of heating and cooling degree days throughout the year for Phoenix in 1980 and 2022. The differences in the two figures demonstrate an increase in the number of cooling days in March, April, and October and a decrease in the number of heating days in February and March.  

```{r number of heating/cooling days Phoenix 1980, fig.cap="Number of Phoenix's Heating and Cooling Degree Days, 1980"}
#Put the heating and cool days on the same graph
plot_Phoenix_1980_HeatCoolDays <- ggplot() +
  geom_bar(data = Phoenix_Cool_1980, aes(x = Month, fill = "Cooling Days")) +
  geom_bar(data = Phoenix_Heat_1980, aes(x = Month, fill = "Heating Days")) +
  scale_x_continuous(breaks = seq_along(month.abb), labels = month.abb) +
  labs(title = "Cooling and Heating Days in Phoenix, 1980", x = "Month",
       y = "Cooling or Heating Days") + 
  scale_fill_manual(values = c("blue", "red"))
print(plot_Phoenix_1980_HeatCoolDays)
```

```{r Phoenix 2022 heat/cool degree days, fig.cap="Number of Phoenix's Heating and Cooling Degree Days, 2022"}
plot_Phoenix_2022_HeatCoolDays <- ggplot() +
  geom_bar(data = Phoenix_Cool_2022, aes(x = Month, fill = "Cooling Days")) +
  geom_bar(data = Phoenix_Heat_2022, aes(x = Month, fill = "Heating Days")) +
  scale_x_continuous(breaks = seq_along(month.abb), labels = month.abb) +
  labs(title = "Cooling and Heating in Phoenix, 2022", x = "Month",
       y = "Cooling or Heating Days") + 
  scale_fill_manual(values = c("blue", "red"))
print(plot_Phoenix_2022_HeatCoolDays)
```

\newpage

Figures 5 and 6 show the intensity of heating and cooling degree days in Phoenix in 1980 and 2022. In this instance, intensity is the number of degrees over or under 65 F per day. The peaks of the heating degree day intensity is approximately half the intensity of the cooling degree day intensity. Figure 5 shows that there were some higher peaks in 2022 of heating degree days, and March of 2022 saw a much lower intensity of heat degree days than March of 1980. Figure 6 shows the intensity of cooling degree days peaking consistently in the month of July. July 2022 reached 39 degrees F and July 1980 reached 36 degrees F. In 2022 compared to 1980, cooling degree days started earlier in the year, and consistently reached a slightly higher intensity.  

```{r code for intensity of cooling/heating degree days, fig.cap="Comparison of Heat Degree Days in Phoenix in 1980 and 2022"}
#monthly labels
month_lab = tibble(doy = yday(as.Date(c("1980-02-15", "1980-04-15",
                                          "1980-06-15", "1980-08-15",
                                          "1980-10-15", "1980-12-15"))),
                     lab = c("Feb", "Apr", "Jun", "Aug", "Oct", "Dec"))

#heating degree days
Phoenix_Heat_2022 <- Phoenix %>%
  group_by(Month) %>%
  filter(Year == "2022") %>%
  select(Date, Year, Month, DayofYear, Heat_Degrees)

Phoenix_Heat_1980 <- Phoenix %>%
  group_by(Month) %>%
  filter(Year == "1980") %>%
  select(Date, Year, Month, DayofYear, Heat_Degrees)

plot_Phoenix_1980_2022_Heat <- ggplot() +
  geom_line(data = Phoenix_Heat_1980, 
            aes(x = DayofYear, y = Heat_Degrees, 
                color = '1980')) +
  geom_line(data = Phoenix_Heat_2022, 
            aes(x = DayofYear, y = Heat_Degrees, 
                color = '2022')) +
  labs(title = "Intensity of Heating Degrees (<65 F) in Phoenix (1980 & 2022)",
       x = "Month", y = "Number of Degrees F below 65 F") +
  scale_x_continuous(breaks = month_lab$doy, labels = month_lab$lab) +
  ylim(0, 40)
print(plot_Phoenix_1980_2022_Heat)

Phoenix_Heat_1980_Lineplot <-
  ggplot(Phoenix_Heat_1980, aes(x = DayofYear, y = Heat_Degrees)) +
  geom_line(color = "darkorange") +
  labs(title = "Intensity of Heating Days (<65 F) in Phoenix (1980)",
       x = "Month", y = "Number of Degrees F below 65 F") +
  scale_x_continuous(breaks = month_lab$doy, labels = month_lab$lab) +
  ylim(0, 40)

Phoenix_Heat_2022_Lineplot <- 
  ggplot(Phoenix_Heat_2022, 
         aes(x = DayofYear, y = Heat_Degrees)) +
  geom_line(color = "purple")+
  labs(title = "Intensity of Heating Days (<65 F) in Phoenix (2022)",
       x = "Month", y = "Number of Degrees F below 65 F") +
  scale_x_continuous(breaks = month_lab$doy, labels = month_lab$lab) +
  ylim(0, 40)

#grid_Phoenix_1980_2022_heat <-
 # plot_grid(Phoenix_Heat_1980_Lineplot, Phoenix_Heat_2022_Lineplot)

```

```{r side by side intensity of heat degree days}
Phoenix_Heat_1980_Lineplot <-
  ggplot(Phoenix_Heat_1980, aes(x = DayofYear, y = Heat_Degrees)) +
  geom_line(color = "darkorange") +
  labs(title = "Intensity of Heating Days (<65 F) in Phoenix (1980)",
       x = "Month", y = "Number of Degrees F below 65 F") +
  scale_x_continuous(breaks = month_lab$doy, labels = month_lab$lab) +
  ylim(0, 40)

Phoenix_Heat_2022_Lineplot <- 
  ggplot(Phoenix_Heat_2022, 
         aes(x = DayofYear, y = Heat_Degrees)) +
  geom_line(color = "purple")+
  labs(title = "Intensity of Heating Days (<65 F) in Phoenix (2022)",
       x = "Month", y = "Number of Degrees F below 65 F") +
  scale_x_continuous(breaks = month_lab$doy, labels = month_lab$lab) +
  ylim(0, 40)

grid_Phoenix_1980_2022_heat <-
  plot_grid(Phoenix_Heat_1980_Lineplot, Phoenix_Heat_2022_Lineplot)
```

```{r Intensity of Heat Degree Days Phoenix 1980 and 2022 }
# Comparing Intensity of Heat Degree Days in Phoenix in 1980 and 2022

plot_Phoenix_1980_2022_Heat <- ggplot() +
  geom_line(data = Phoenix_Heat_1980, 
            aes(x = DayofYear, y = Heat_Degrees),
                color = 'darkorange') +
  geom_line(data = Phoenix_Heat_2022, 
            aes(x = DayofYear, y = Heat_Degrees), 
                color = 'purple') +
  labs(title = "Intensity of Heating Degrees (<65 F) in Phoenix (1980 & 2022)",
       x = "Month", y = "Number of Degrees F below 65 F") +
  scale_x_continuous(breaks = month_lab$doy, labels = month_lab$lab) +
  ylim(0, 40)
#print(plot_Phoenix_1980_2022_Heat)
```

```{r cooling degree days intensity, fig.cap="Comparison of Cool Degree Days in Phoenix in 1980 and 2022"}
#cooling degree days
Phoenix_Cooling_2022 <- Phoenix %>%
  group_by(Month) %>%
  filter(Year == "2022") %>%
  select(Date, Year, Month, DayofYear, Cool_Degrees)

Phoenix_Cooling_1980 <- Phoenix %>%
  group_by(Month) %>%
  filter(Year == "1980") %>%
  select(Date, Year, Month, DayofYear, Cool_Degrees)

plot_Phoenix_1980_2022_Cool <- ggplot() +
  geom_line(data = Phoenix_Cooling_1980, 
            aes(x = DayofYear, y = Cool_Degrees, 
            color='1980')) +
  geom_line(data = Phoenix_Cooling_2022, 
            aes(x = DayofYear, y = Cool_Degrees,
            color = '2022')) +
  labs(title = "Intensity of Cooling Degrees (>65 F) in Phoenix (1980 & 2022)",
       x = "Month", y = "Number of Degrees F above 65 F") +
  scale_x_continuous(breaks = month_lab$doy, labels = month_lab$lab) +
  ylim(0, 40)
print(plot_Phoenix_1980_2022_Cool)
```

\newpage

# 4. Analysis

## Question 1: Has there been an increase in the number of heating and cooling days in the U.S. Southwest from 1980 to 2022?

We used a time series analysis for the heating and cooling degree days for each of the three major cities to address this research question. We made univariate time series objects for the heating and cooling degree days for each city, and then decomposed each time series object to separate out the seasonality, trend, and remainder from the data. The graphs below showcase the time series plots.

```{r creating time series objects for heat and cool degrees}
#Creating univariate time series objects for heat and cool degrees for each major city

#Phoenix
Phoenix_heatdegrees_ts <- ts(Phoenix$Heat_Degrees, 
                            start = c(1980, 1),
                            frequency = 365)

Phoenix_cooldegrees_ts <- ts(Phoenix$Cool_Degrees, 
                            start = c(1980, 1),
                            frequency = 365)

#Las Vegas
LasVegas_heatdegrees_ts <- ts(LasVegas$Heat_Degrees,
                              start = c(1980, 1),
                              frequency = 365)

LasVegas_cooldegrees_ts <- ts(LasVegas$Cool_Degrees,
                              start = c(1980, 1),
                              frequency = 365)

#San Diego
SanDiego_heatdegrees_ts <- ts(SanDiego$Heat_Degrees,
                              start = c(1980, 1),
                              frequency = 365)

SanDiego_cooldegrees_ts <- ts(SanDiego$Cool_Degrees,
                              start = c(1980, 1),
                              frequency = 365)
```

```{r decomposing for heat/cool degrees, out.width="50%"}
# Decomposing each ts object

Phoenix_heatdegrees_ts_Decomposed <-
  stl(Phoenix_heatdegrees_ts, s.window = "periodic")

plot(Phoenix_heatdegrees_ts_Decomposed, main="Phoenix Heat Degree Days")

Phoenix_cooldegrees_ts_Decomposed <-
  stl(Phoenix_cooldegrees_ts, s.window = "periodic")

plot(Phoenix_cooldegrees_ts_Decomposed, main="Phoenix Cool Degree Days")

#Las Vegas decomposition and plot

LasVegas_heatdegrees_ts_Decomposed <-
  stl(LasVegas_heatdegrees_ts, s.window = "periodic")

plot(LasVegas_heatdegrees_ts_Decomposed, main="Las Vegas Heat Degree Days")

LasVegas_cooldegrees_ts_Decomposed <-
  stl(LasVegas_cooldegrees_ts, s.window = "periodic")

plot(LasVegas_cooldegrees_ts_Decomposed, main = "Las Vegas Cool Degree Days")

# San Diego decomposition and plot

SanDiego_heatdegrees_ts_Decomposed <-
  stl(SanDiego_heatdegrees_ts, s.window = "periodic")

plot(SanDiego_heatdegrees_ts_Decomposed, main = "San Diego Heat Degree Days")

SanDiego_cooldegrees_ts_Decomposed <-
  stl(SanDiego_cooldegrees_ts, s.window = "periodic")

plot(SanDiego_cooldegrees_ts_Decomposed, main = "San Diego Cool Degree Days")

```

The Seasonal Mann Kendall test was performed on each of the city's heating and cooling degree days data to analyze if the data is stationary. The null hypothesis for the Seasonal Mann Kendall test is that the data is stationary, meaning there is no significant trend over time. The alternative hypothesis is that there is a significant trend in the data over time.

The results of the Seasonal Mann Kendall test for the heating degree days in Phoenix, Las Vegas, and San Diego are found in Table 1. The results indicate that Phoenix and Las Vegas have a p-value of less than 0.05 (3.0198e-13 and 2.22e-16, respectively), so the null hypothesis can be rejected. The p value for San Diego is greater than 0.5 (having a value of 0.49529), so we cannot reject the null hypothesis. The tau value indicates the slope of the trend, and all are negative, but this is only significant for only Phoenix and Las Vegas. Las Vegas has the steepest slope. 

```{r SMK test for heat degree days, message=FALSE, warning=FALSE, include=FALSE}
# Run Seasonal Mann Kendall Test for heating degree days

#Phoenix
Phoenix_heatdegrees_trend <- SeasonalMannKendall(Phoenix_heatdegrees_ts)
Phoenix_heatdegrees_trend
# tau = -0.0592 and 2-sided p value =3.0198e-13 (REJECT NULL)

#Las Vegas
LasVegas_heatdegrees_trend <- SeasonalMannKendall(LasVegas_heatdegrees_ts)
LasVegas_heatdegrees_trend
# tau = -0.16, 2-sided p value =< 2.22e-16 (REJECT NULL)

#San Diego
SanDiego_heatdegrees_trend <- SeasonalMannKendall(SanDiego_heatdegrees_ts)
SanDiego_heatdegrees_trend
# tau = -0.00414, 2-sided pvalue = 0.49529 (CANNOT REJECT NULL)
```

```{r heating degree day kable}
heatdegree_SMK_cities <- c("Phoenix", "Las Vegas", "San Diego")

heatdegree_SMK_p_values <- c("3.0198e-13", "2.22e-16", "0.49529")

heatdegree_SMK_tau_values <- c("-0.0592", "-0.16", "-0.00414")

heatdegree_SMK <- data.frame(heatdegree_SMK_cities, heatdegree_SMK_p_values, 
                             heatdegree_SMK_tau_values)

kable(heatdegree_SMK, "pipe", col.names =  c("City", "p-value", "tau value"), 
caption = "Results for the Seasonal Mann Kendall test for Heating Degree Days", align = c("l", "c", "c"), 
digits = 4, scientific = TRUE)
```

Table 2 shows the results of the Seasonal Mann Kendall Test for the cooling degree days over time in Phoenix, Las Vegas, and San Diego. Phoenix and Las Vegas have a p value of less than 0.05 (p value = 2.22e-16 for each city), so the null hypothesis can be rejected. The heating degree days for Phoenix and Las Vegas have a significant trend over time. The cooling degree days for San Diego produced a p value of 0.36317, which is higher the alpha level of 0.05, so we cannot reject the null hypothesis for this location. There is not a significant trend over time in the cooling degree days for San Diego. The tau values are the slope of the trend line, and indicate if the trend is positive or negative. For both Phoenix and Las Vegas, the slope is positive, meaning there is a positive trend over time in cooling degree days in these locations. 

```{r SMK test for cool degree days, message=FALSE, warning=FALSE, include=FALSE}
# Run Seasonal Mann Kendall Test for cooling degree days

Phoenix_cooldegrees_trend <-SeasonalMannKendall(Phoenix_cooldegrees_ts)
Phoenix_cooldegrees_trend
#p value =< 2.22e-16 ; reject the null hypothesis (data is stationary)
# tau = 0.0983 (slope of trend)

#Las Vegas
LasVegas_cooldegrees_trend <-SeasonalMannKendall(LasVegas_cooldegrees_ts)
LasVegas_cooldegrees_trend
# p value is =< 2.22e-16 ; reject the null hypothesis (data is stationary) 
# tau = 0.189 (slope of the trend)

#San Diego
SanDiego_cooldegrees_trend <- SeasonalMannKendall(SanDiego_cooldegrees_ts)
SanDiego_cooldegrees_trend
# p value is 0.36317 ; CANNOT reject the null 
# tau = -0.00542
```

```{r cooling degree day kable}
cooldegree_SMK_cities <- c("Phoenix", "Las Vegas", "San Diego")

cooldegree_SMK_p_values <- c("2.22e-16", "2.22e-16", "0.36317")

cooldegree_SMK_tau_values <- c("0.0983", "0.189", "-0.00542")

cooldegree_SMK <- data.frame(cooldegree_SMK_cities, cooldegree_SMK_p_values, 
                             cooldegree_SMK_tau_values)

kable(cooldegree_SMK, "pipe", col.names =  c("City", "p-value", "tau value"), 
caption = "Results for the Seasonal Mann Kendall test for Cooling Degree Days", align = c("l", "c", "c"), 
digits = 4, scientific = TRUE)
```

In summary, we can conclude there has been a significant increase in the trend of cooling degree days over time in some cities in the U.S. Southwest from 1980 to 2022, but not uniformly. There has been some significant decrease in the trend of heating degree days over time in some U.S. Southwest cities from 1980 to 2022, but not in all locations. The increase in cooling degree days and decrease in heating degree days is evidence of the overall warming temperatures in Las Vegas and Phoenix. Tests performed on the San Diego dataset indicated that this data may be stationary, meaning there is no significant trend upward or downward in the heating and cooling degree days over time.

\newpage

## Question 2: Is there a difference in heating and cooling days in urban and rural spaces in the U.S. Southwest over time?

## Question 3: How have minimum and maximum temperatures changed over time?
To address the question of whether minimum and maximum temperatures have changed over time, we conducted a time series analysis on every city. We first created univariate time series objects for each measure for each site. We then decomposed each time series to review the trends and seasonality present.
```{r time series objects for min and max temperatures}
#Creating univariate time series objects for min & max temperature for each site
#Pheonix
Phoenix_temp_min_ts <- ts(Phoenix$Daily_Temp_Min,
                            start = c(1980, 1),
                            frequency = 365)

Phoenix_temp_max_ts <- ts(Phoenix$Daily_Temp_Max, 
                            start = c(1980, 1),
                            frequency = 365)

#Fountain Hills
FountainHills_temp_min_ts <- ts(FountainHills$Daily_Temp_Min,
                            start = c(1980, 1),
                            frequency = 365)

FountainHills_temp_max_ts <- ts(FountainHills$Daily_Temp_Max, 
                            start = c(1980, 1),
                            frequency = 365)

#Las Vegas
LasVegas_temp_min_ts <- ts(LasVegas$Daily_Temp_Min,
                            start = c(1972, 1),
                            frequency = 365)

LasVegas_temp_max_ts <- ts(LasVegas$Daily_Temp_Max, 
                            start = c(1972, 1),
                            frequency = 365)

#Pahrump
Pahrump_temp_min_ts <- ts(Pahrump$Daily_Temp_Min,
                            start = c(1972, 1),
                            frequency = 365)

Pahrump_temp_max_ts <- ts(Pahrump$Daily_Temp_Max, 
                            start = c(1972, 1),
                            frequency = 365)

#San Diego
SanDiego_temp_min_ts <- ts(SanDiego$Daily_Temp_Min,
                            start = c(1972, 1),
                            frequency = 365)

SanDiego_temp_max_ts <- ts(SanDiego$Daily_Temp_Max, 
                            start = c(1972, 1),
                            frequency = 365)

#Ramona
Ramona_temp_min_ts <- ts(Ramona$Daily_Temp_Min,
                            start = c(1998, 1),
                            frequency = 365)

Ramona_temp_max_ts <- ts(Ramona$Daily_Temp_Max, 
                            start = c(1998, 1),
                            frequency = 365)
```

```{r decomposing time series objects for min and max temperatures, out.width="50%"}
# Decompose each time series (ts) object
#Phoenix
Phoenix_temp_max_ts_Decomposed <- stl(Phoenix_temp_max_ts,
                                      s.window = "periodic")
plot(Phoenix_temp_max_ts_Decomposed, main = "Phoenix Max Temperature")

Phoenix_temp_min_ts_Decomposed <- stl(Phoenix_temp_min_ts, 
                                      s.window = "periodic")
plot(Phoenix_temp_min_ts_Decomposed, main = "Phoenix Min Temperature")

#Fountain Hills
FountainHills_temp_max_ts_Decomposed <- stl(FountainHills_temp_max_ts,
                                            s.window = "periodic")
plot(FountainHills_temp_max_ts_Decomposed, main = "Fountain Hills Max Temperature")

FountainHills_temp_min_ts_Decomposed <- stl(FountainHills_temp_min_ts,
                                            s.window = "periodic")
plot(FountainHills_temp_min_ts_Decomposed, main = "Fountain Hills Min Temperature")

#Las Vegas
LasVegas_temp_max_ts_Decomposed <- stl(LasVegas_temp_max_ts,
                                       s.window = "periodic")
plot(LasVegas_temp_max_ts_Decomposed, main = "Las Vegas Max Temperature")

LasVegas_temp_min_ts_Decomposed <- stl(LasVegas_temp_min_ts,
                                       s.window = "periodic")
plot(LasVegas_temp_min_ts_Decomposed, main = "Las Vegas Min Temperature")

#Pahrump
Pahrump_temp_max_ts_Decomposed <- stl(Pahrump_temp_max_ts,
                                      s.window = "periodic")
plot(Pahrump_temp_max_ts_Decomposed, main = "Pahrump Max Temperature")

Pahrump_temp_min_ts_Decomposed <- stl(Pahrump_temp_min_ts,
                                      s.window = "periodic")
plot(Pahrump_temp_min_ts_Decomposed, main = "Pahrump Min Temperature")

#San Diego
SanDiego_temp_max_ts_Decomposed <- stl(SanDiego_temp_max_ts,
                                       s.window = "periodic")
plot(SanDiego_temp_max_ts_Decomposed, main = "San Diego Max Temperature")

SanDiego_temp_min_ts_Decomposed <- stl(SanDiego_temp_min_ts,
                                       s.window = "periodic")
plot(SanDiego_temp_min_ts_Decomposed, main = "San Diego Min Temperature")

#Ramona
Ramona_temp_max_ts_Decomposed <- stl(Ramona_temp_max_ts,
                                     s.window = "periodic")
plot(Ramona_temp_max_ts_Decomposed, main = "Ramona Max Temperature")

Ramona_temp_min_ts_Decomposed <- stl(Ramona_temp_min_ts,
                                    s.window = "periodic")
plot(Ramona_temp_min_ts_Decomposed, main = "Ramona Min Temperature")
```

We then conducted a Seasonal Mann  Kendall test to test whether the hypothesis that the data is stationary. This would mean that the data exhibit no statistically-significant upward or downward trend over time, meaning that there is no change in the temperature minimum and maximums over time.
```{r SMK tests for min and max temperatures}
#Run Seasonal Mann Kendall Tests

#Phoenix
Phoenix_temp_max_trend <- SeasonalMannKendall(Phoenix_temp_max_ts)
Phoenix_temp_min_trend <- SeasonalMannKendall(Phoenix_temp_min_ts)

#Fountain Hills
FountainHills_temp_max_trend <- SeasonalMannKendall(FountainHills_temp_max_ts)
FountainHills_temp_min_trend <- SeasonalMannKendall(FountainHills_temp_min_ts)

#Las Vegas
LasVegas_temp_max_trend <- SeasonalMannKendall(LasVegas_temp_max_ts)
LasVegas_temp_min_trend <- SeasonalMannKendall(LasVegas_temp_min_ts)

#Pahrump
Pahrump_temp_max_trend <- SeasonalMannKendall(Pahrump_temp_max_ts)
Pahrump_temp_min_trend <- SeasonalMannKendall(Pahrump_temp_min_ts)

#San Diego
SanDiego_temp_max_trend <- SeasonalMannKendall(SanDiego_temp_max_ts)
SanDiego_temp_min_trend <- SeasonalMannKendall(SanDiego_temp_min_ts)

#Ramona
Ramona_temp_max_trend <- SeasonalMannKendall(Ramona_temp_max_ts)
Ramona_temp_min_trend <- SeasonalMannKendall(Ramona_temp_min_ts)
```
The results of the Seasonal Mann Kendall test show us that all trends are statistically significant, so none of the data are stationary. The tau value shows the slope of the trend. Interestingly, Fountain Hill's maximum temperatures, San Diego's minimum temperatures, and Ramona's maximum and minimum temperatures have a downward trend. Every other variable has an upward trend.
``` {r ktable with results from the Seasonal Mann Kendall test}
temp_SMK_cities <- c("Phoenix", "Phoenix", "Fountain Hills", "Fountain Hills",
            "Las Vegas", "Las Vegas", "Pahrump", "Pahrump",
            "San Diego", "San Diego", "Ramona", "Ramona")

temp_SMK_variable <- c("max temperature", "min temperature", "max temperature",
                   "min temperature", "max temperature", "min temperature",
                   "max temperature", "min temperature", "max temperature",
                   "min temperature", "max temperature", "min temperature")

temp_SMK_p_values <- c("<2.22 e-16", "<2.22 e-16", " 7.66 e-07", "<2.22 e-16", 
                  "<2.22 e-16", "<2.22 e-16", "<2.22 e-16", "<2.22 e-16", 
                  "8.91 e-4", "5.10 e-4", "<2.22 e-16", "<2.22 e-16")

temp_SMK_tau_values <- c("0.0617", "0.0951", "-0.0282", "0.0979",
                    "0.0671", "0.29", "0.0482", "0.0678",
                    "0.0173",  "-0.0182",  "-0.112",  "-0.0912")

temp_SMK <- data.frame(temp_SMK_cities, temp_SMK_variable,
                       temp_SMK_p_values, temp_SMK_tau_values)

kable(temp_SMK, "pipe", col.names =  c("City", "Variable", "p-value", "tau value"), 
caption = "Results for the Seasonal Mann Kendall test on minimum and maximum
temperatures", align = c("l", "c", "c", "c"), digits = 4, scientific = TRUE)
```

``` {r plots for min and max temperatures, include = FALSE}
#Phoenix plot
Phoenix_plot_temp <- ggplot(Phoenix, aes(x = Date)) +
  geom_line(aes(y = Daily_Temp_Max, color = "Max Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Max), method = lm, color = "red") + 
  geom_line(aes(y = Daily_Temp_Min, color = "Min Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Min), method = lm, color = "blue") + 
  scale_color_manual(values = c("pink", "lightblue")) +
  labs(title = "Phoenix (1980-2022)", x = "Year",
       y = "Daily Temp Peaks (°F)", color = "Legend")
print(Phoenix_plot_temp)

#Fountain Hills plot
FountainHills_plot_temp <- ggplot(FountainHills, aes(x = Date)) +
  geom_line(aes(y = Daily_Temp_Max, color = "Max Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Max), method = lm, color = "red") + 
  geom_line(aes(y = Daily_Temp_Min, color = "Min Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Min), method = lm, color = "blue") + 
  scale_color_manual(values = c("pink", "lightblue")) +
  labs(title = "Fountain Hills (1980-2022)", x = "Year",
       y = "Daily Temp Peaks (°F)", color = "Legend")
print(FountainHills_plot_temp)

#Las Vegas
LasVegas_plot_temp <- ggplot(LasVegas, aes(x = Date)) +
  geom_line(aes(y = Daily_Temp_Max, color = "Max Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Max), method = lm, color = "red") + 
  geom_line(aes(y = Daily_Temp_Min, color = "Min Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Min), method = lm, color = "blue") + 
  scale_color_manual(values = c("pink", "lightblue")) +
  labs(title = "Las Vegas (1972-2022)", x = "Year",
       y = "Daily Temp Peaks (°F)", color = "Legend")
print(LasVegas_plot_temp)

#Pahrump
Pahrump_plot_temp <- ggplot(Pahrump, aes(x = Date)) +
  geom_line(aes(y = Daily_Temp_Max, color = "Max Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Max), method = lm, color = "red") + 
  geom_line(aes(y = Daily_Temp_Min, color = "Min Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Min), method = lm, color = "blue") + 
  scale_color_manual(values = c("pink", "lightblue")) +
  labs(title = "Pahrump (1972-2022)", x = "Year",
       y = "Daily Temp Peaks (°F)", color = "Legend")
print(Pahrump_plot_temp)

#San Diego
SanDiego_plot_temp <- ggplot(SanDiego, aes(x = Date)) +
  geom_line(aes(y = Daily_Temp_Max, color = "Max Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Max), method = lm, color = "red") + 
  geom_line(aes(y = Daily_Temp_Min, color = "Min Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Min), method = lm, color = "blue") + 
  scale_color_manual(values = c("pink", "lightblue")) +
  labs(title = "San Diego (1972-2022)", x = "Year",
       y = "Daily Temp Peaks (°F)", color = "Legend")
print(SanDiego_plot_temp)

#Ramona
Ramona_plot_temp <- ggplot(Ramona, aes(x = Date)) +
  geom_line(aes(y = Daily_Temp_Max, color = "Max Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Max), method = lm, color = "red") + 
  geom_line(aes(y = Daily_Temp_Min, color = "Min Temp")) + 
  geom_smooth(aes(y = Daily_Temp_Min), method = lm, color = "blue") + 
  scale_color_manual(values = c("pink", "lightblue")) +
  labs(title = "Ramona (1998-2022)", x = "Year",
       y = "Daily Temp Peaks (°F)", color = "Legend")
print(Ramona_plot_temp)
```

``` {r put above plots in a single grid,  fig.width = 7.5, fig.height = 10}
#create a title
title <- ggdraw() + draw_label("Trends in Minimum and Maximum Temperatures across Cities")

#extract a legend from one of the plots, make it into an object
plot_legend <- get_legend(Ramona_plot_temp)

#add the plots without legends to a cowplot
temp_plots <- plot_grid(
  Phoenix_plot_temp + theme(legend.position = "none"),
  FountainHills_plot_temp + theme(legend.position = "none"),
  LasVegas_plot_temp + theme(legend.position = "none"),
  Pahrump_plot_temp + theme(legend.position = "none"),
  SanDiego_plot_temp + theme(legend.position = "none"),
  Ramona_plot_temp + theme(legend.position = "none"),
  align = c("hv"), ncol = 2, nrow = 3)

#combine plots and legend
temperature_plots <- plot_grid(title, plot_legend, temp_plots, ncol = 1, nrow = 3, rel_heights = c(0.5, 0.5, 10))
print(temperature_plots)
```

The graphs above show the trends increasing or slightly decreasing in the case of .

\newpage

# 5. Summary and Conclusions


\newpage

# 6. References

US EPA, O. (2016, July 1). Climate Change Indicators: Heating and Cooling Degree Days [Reports and Assessments]. https://www.epa.gov/climate-indicators/climate-change-indicators-heating-and-cooling-degree-days
